Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2023-01-11 src/docs_paper/paper.Rmd
Iterate paper and appendix

---
title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A
    sepNext: ","
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
  - name: Seth R.
    surname: Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    label: e3
    addressLabel: C
    sepNext: ","
  - name: Jeffrey W.
    surname: Eaton
    email: jeffrey.eaton@imperial.ac.uk
    label: e4
    addressLabel: D
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
  - label: C
    name: Department of Computer Science, University of Oxford
    authorsLabels: e3
  - label: D
    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
    authorsLabels: e4
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach used currently, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler. By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - spatial statistics
  - small-area estimation
  - INLA
  - AGHQ
  - HIV epidemiology

@@

Situational requirements combined with the complexity of the model present a difficult Bayesian inference problem.
Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
Owing to its speed and flexibility, `TMB` has recently been gaining popularity spatial statistics [@osgoodzimmerman2021statistical].
Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
For the Naomi model, we use this option to integrate out the latent field parameters.
Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
As such, we have been motivated to find a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.

@@

@eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
The model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
We consider a simplified version of the model which is defined only at $T_1$, omitting temporal projection to $T_2$ and $T_3$.
Below we provide an overview of the three components in the simplified model, highlighting aspects which make it a challenge for existing inferential approaches.
More complete mathematical details of this simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.

@@

Consider a country in sub-Saharan Africa where a household survey has taken place at $T_1$.
Let $x$ index district, $a$ index five-year age band, $s$ index sex and $t$ index time.
For ease of notation, let $i$ index the coarsest district-age-sex division included in the model.
The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
Let $N_i$ be the population size, and consider the following three indicators: HIV prevalence $\rho_i \in [0, 1]$, ART coverage $\alpha_i \in [0, 1]$, and annual HIV incidence rate $\lambda_i > 0$.

We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale
$$
\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}),
$$
where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$.

Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes 
$$
M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
where $j$ index individuals in all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
The observed number of indicator cases is then
$$
Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
$$
For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood
$$
Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}),
$$
where $\theta_{\mathcal{I}}$ are the following weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}
We link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ by
$$
\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where $\Omega_T$ is the mean duration of recent infection and $\beta_T$ is the false recent ratio.

@@

We model HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women as being offset on the logit scale from the general population indicator as follows
\begin{align*}
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \tilde{\eta}^{\rho^\text{ANC}}_i, \\
\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \tilde{\eta}^{\rho^\text{ANC}}_i.
\end{align*}

Let $W^\text{ANC}_i$ be the number of ANC clients, $X^\text{ANC}_i$ the number of those with ascertained status, $Y^\text{ANC}_i$ the number of those with positive status, and $Z^\text{ANC}_i$ the number of ANC clients already on ART prior to their first ANC visit.
ANC data from the year of the most recent survey is included in the model.

For women 15-49 the predicted number of ANC clients $\Psi_i$ is a log-linear model
$$
\log (\Psi_i) = \log (N_i) + \tilde \eta^\Psi
$$

@@

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.

Let $\dot A_i$ be the number of people receiving ART.
Likelihood section to be written.

>>>>

---
title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A
    sepNext: ","
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
  - name: Seth R.
    surname: Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    label: e3
    addressLabel: C
    sepNext: ","
  - name: Jeffrey W.
    surname: Eaton
    email: jeffrey.eaton@imperial.ac.uk
    label: e4
    addressLabel: D
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
  - label: C
    name: Department of Computer Science, University of Oxford
    authorsLabels: e3
  - label: D
    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
    authorsLabels: e4
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach which is currently in use, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler. By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - spatial statistics
  - small-area estimation
  - INLA
  - AGHQ
  - HIV epidemiology

@@

The use case requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
Owing to its speed and flexibility, `TMB` has recently been gaining popularity spatial statistics [@osgoodzimmerman2021statistical].
Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
For the Naomi model, we use this option to integrate out the latent field parameters.
Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
As such, we have been motivated to find a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.

@@

@eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
The model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
We consider a simplified version of the model which is defined only at $T_1$ omitting temporal projection to $T_2$ and $T_3$.
Below we provide an overview of the three components in the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.

@@

Consider a country in sub-Saharan Africa where a household survey has taken place during the quarter $T_1$.
Let $x$ index district, $a$ five-year age band, $s$ sex and $t$ time.
For ease of notation, let $i$ index the coarsest district-age-sex division included in the model.
The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be ART coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale
$$
\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}),
$$
where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes 
$$
M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
where $j$ indexes individuals across all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
The observed number of indicator cases is then
$$
Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
$$
For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood
$$
Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}),
$$
where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}
To link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ we use
$$
\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where the mean duration of recent infection $\Omega_T$ and  is the false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.

@@

We model HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women as being offset on the logit scale from the general population indicator as follows
\begin{align*}
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\rho^\text{ANC}}_i.
\end{align*}
We inform these processes using the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $X^\text{ANC}_\mathcal{I}$, the number of those with positive status $Y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $Z^\text{ANC}_\mathcal{I}$.
We use the binomial working likelihoods
\begin{align*}
Y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(Y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}) \\
Z^\text{ANC}_\mathcal{I} &\sim \text{Bin}(Y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}),
\end{align*}
where again we use weighted aggregates
\begin{align*}
\rho^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in \mathcal{I}} \Psi_i}, \\
\alpha^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}},
\end{align*}
with $\Psi_i$ the number of pregnant women.

@@

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$.

Let $\dot A_\mathcal{I}$ be the number of people receiving ART.
